Long-term care homes were hit hardest by 2023 disease outbreaks in Toronto healthcare facilities

### 2024-01-22 outputs/paper/paper.qmd
Finished the Abstract and Introduction sections of the paper

---
title: "My title"
author: 
  - Benny Rochwerg
thanks: "A GitHub Repository containing all data, R code, and other files used in this investigation is located here: <https://github.com/bennyrochwerg/healthcare-outbreaks-toronto>"
date: 01-23-2024
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

@@

```{r}
#| echo: false
#| warning: false
#| label: tbl-outbreaks-by-location-type
#| tbl-cap: "Outbreaks by type of healthcare location in Toronto in 2023"
#| tbl-colwidths: [50, 20, 20]

@@

@tbl-outbreaks-by-location-type and @fig-outbreaks-by-location-type illustrate that long-term care homes comprise a majority (approximately 56\%) of the total number of healthcare outbreaks in Toronto. This is followed by retirement homes (approximately 18\%), chronic care settings in hospitals (approximately 12\%), acute care settings in hospitals (approximately 10\%), psychiatric care settings in hospitals (approximately 2\%), and transitional care locations (approximately 1\%). Note that these percentages do not add up to 100\% as the proportions in @tbl-outbreaks-by-location-type were rounded to two decimal places.

```{r}
#| echo: false
#| warning: false
#| label: tbl-outbreaks-by-outbreak-type
#| tbl-cap: "Outbreaks by type of outbreak for Toronto healthcare locations in 2023"
#| tbl-colwidths: [50, 20, 20]

@@

@tbl-outbreaks-by-outbreak-type and @fig-outbreaks-by-outbreak-type show that the vast majority of healthcare outbreaks in Toronto are respiratory in nature (approximately 95\%). This is followed by enteric outbreaks (approximately 4\%) and other types of outbreaks (less than 1\%). Note that these percentages do not add up to 100\% as the proportions in @tbl-outbreaks-by-outbreak-type were rounded to two decimal places.

```{r}
#| echo: false
#| warning: false
#| label: tbl-outbreaks-by-first-known-cause
#| tbl-cap: "Outbreaks by first known cause for Toronto healthcare locations in 2023"
#| tbl-colwidths: [50, 20, 20]

@@

@tbl-outbreaks-by-first-known-cause and @fig-outbreaks-by-first-known-cause illustrate that the majority of Toronto healthcare outbreaks have COVID-19 as their first known cause (approximately 66\%). This is followed by rhinovirus (approximately 7\%), respiratory syncytial virus (approximately 5\%), parainfluenza (approximately 3\%), influenza A (approximately 3\%), seasonal coronavirus (approximately 3\%), metapneumovirus (approximately 2\%), norovirus (approximately 2\%), enterovirus/rhinovirus (approximately 1\%), enterovirus (approximately 1\%), CPE (less than 1\%), neonatal Group B streptococcal disease (less than 1\%), and streptococcus pyogenes (less than 1\%). Approximately 7\% of outbreaks had a pending or unknown first known cause. Note that these percentages do not add up to 100\% as the proportions in @tbl-outbreaks-by-first-known-cause were rounded to two decimal places.

@@

# Introduction

You can and should cross-reference sections and sub-sections.

The remainder of this paper is structured as follows. @sec-data....

## Discussion {#data-discussion}

In @sec-data-results, it was observed that long-term care homes had the greatest proportion of total outbreaks among healthcare facilities in Toronto in 2023 (@tbl-outbreaks-by-location-type and @fig-outbreaks-by-location-type), that approximately 95\% of outbreaks were respiratory in nature (@tbl-outbreaks-by-outbreak-type and @fig-outbreaks-by-outbreak-type), and that COVID-19 was the first known cause of a majority of outbreaks (@tbl-outbreaks-by-first-known-cause and @fig-outbreaks-by-first-known-cause).

Future research should be conducted to assess the state of disease outbreaks in long-term care homes in particular, both in Toronto and across Canada, in order to gain deeper insight into this critical issue. The prevalence of disease outbreaks in other settings, such as schools and workplaces, should also be investigated to expand on this analysis.

>>>>

---
title: "Long-term care homes were hit hardest by 2023 disease outbreaks in Toronto healthcare facilities"
author: 
  - Benny Rochwerg
thanks: "A GitHub Repository containing all data, R code, and other files used in this investigation is located here: <https://github.com/bennyrochwerg/healthcare-outbreaks-toronto>"
date: 01-23-2024
date-format: long
abstract: "In this report, Toronto Public Health data on 2023 outbreaks in healthcare facilities was analyzed. Long-term care homes comprised a majority of the outbreaks, approximately 95% of the outbreaks were respiratory, and almost two-thirds of the outbreaks had COVID-19 as their first known cause. This highlights the need for enhanced public health measures in healthcare facilities, particularly in long-term care homes. Future research should assess disease outbreaks in additional settings, such as workplaces and schools, to gain further insight into the extent of this issue."
format: pdf
number-sections: true
# The toc parameter used to create a table of contents was obtained from
# Quarto documentation:
# https://quarto.org/docs/reference/formats/pdf.html
toc: true
# The tbl-colwidths parameter used to adjust the table column widths was
# obtained from Quarto documentation:
# https://quarto.org/docs/authoring/tables.html
tbl-colwidths: [50, 20, 20]
bibliography: references.bib
---

@@

In March 2020, the World Health Organization proclaimed that the spread of COVID-19 constituted a pandemic [@whocovid]. Shortly thereafter, in May 2020, the COVID-19 mortality rate in Canadian long-term care homes reached approximately 35\% and represented 81\% of Canadian COVID-19 deaths [@cihi]. In general, the risk of disease circulation is greater in healthcare settings due to an increased presence of ill individuals and nearer physical contact between patients and staff [@cdc].

Consequently, with the recent advent of large volumes of respiratory diseases such as influenza and COVID-19 in Toronto [@covidflu], assessing the prevalence of disease outbreaks in healthcare facilities in Toronto is of paramount importance. As a result, this investigation examines the number of outbreaks by location type (long-term care homes, retirement homes, hospitals, and transitional care facilities), outbreak type (respiratory and enteric/gastroenteric, the latter referring to intestinal and/or stomach illness [@enteric; @gastroenteric]), and first known cause (e.g., COVID-19, influenza A, etc.) to fill this knowledge gap.

To do this, Toronto Public Health data on 2023 outbreaks in healthcare facilities was first obtained as described in @sec-data-overview. Based on this dataset, it was discovered that long-term care homes comprised a majority of the outbreaks, that approximately 95\% of the outbreaks were respiratory, and that almost two-thirds of the outbreaks had COVID-19 as their first known cause (@sec-data-results). As discussed in @sec-data-discussion, the high prevalence of outbreaks in long-term care homes has likely been exacerbated by the removal of a universal masking requirement, suggesting that the Government of Ontario should consider reinstating this policy to protect these vulnerable individuals. With respect to the structure of this paper, @sec-data includes an overview of the data, results, and discussion; @sec-acknowledgements acknowledges a valuable source of knowledge in creating this report; and @sec-appendix includes supplementary information.

@@

```{r}
#| echo: false
#| warning: false
#| label: tbl-outbreaks-by-location-type
#| tbl-cap: "Outbreaks by type of healthcare location in Toronto in 2023"

@@

@tbl-outbreaks-by-location-type and @fig-outbreaks-by-location-type illustrate that long-term care homes comprised a majority (approximately 56\%) of the total number of 2023 healthcare outbreaks in Toronto. This was followed by retirement homes (approximately 18\%), chronic care settings in hospitals (approximately 12\%), acute care settings in hospitals (approximately 10\%), psychiatric care settings in hospitals (approximately 2\%), and transitional care locations (approximately 1\%). Note that these percentages do not add up to 100\% as the proportions in @tbl-outbreaks-by-location-type were rounded to two decimal places.

```{r}
#| echo: false
#| warning: false
#| label: tbl-outbreaks-by-outbreak-type
#| tbl-cap: "Outbreaks by type of outbreak for Toronto healthcare locations in 2023"

@@

@tbl-outbreaks-by-outbreak-type and @fig-outbreaks-by-outbreak-type show that the vast majority of 2023 healthcare outbreaks in Toronto were respiratory in nature (approximately 95\%). This was followed by enteric outbreaks (approximately 4\%) and other types of outbreaks (less than 1\%). Note that these percentages do not add up to 100\% as the proportions in @tbl-outbreaks-by-outbreak-type were rounded to two decimal places.

```{r}
#| echo: false
#| warning: false
#| label: tbl-outbreaks-by-first-known-cause
#| tbl-cap: "Outbreaks by first known cause for Toronto healthcare locations in 2023"

@@

@tbl-outbreaks-by-first-known-cause and @fig-outbreaks-by-first-known-cause illustrate that the majority of 2023 Toronto healthcare outbreaks had COVID-19 as their first known cause (approximately 66\%). This was followed by rhinovirus (approximately 7\%), respiratory syncytial virus (approximately 5\%), parainfluenza (approximately 3\%), influenza A (approximately 3\%), seasonal coronavirus (approximately 3\%), metapneumovirus (approximately 2\%), norovirus (approximately 2\%), enterovirus/rhinovirus (approximately 1\%), enterovirus (approximately 1\%), CPE (less than 1\%), neonatal Group B streptococcal disease (less than 1\%), and streptococcus pyogenes (less than 1\%). Approximately 7\% of outbreaks had a pending or unknown first known cause. Note that these percentages do not add up to 100\% as the proportions in @tbl-outbreaks-by-first-known-cause were rounded to two decimal places.

## Discussion {#sec-data-discussion}

In @sec-data-results, it was observed that long-term care homes had the greatest proportion of the total outbreaks among healthcare facilities in Toronto in 2023 (@tbl-outbreaks-by-location-type and @fig-outbreaks-by-location-type), that approximately 95\% of the outbreaks were respiratory in nature (@tbl-outbreaks-by-outbreak-type and @fig-outbreaks-by-outbreak-type), and that COVID-19 was the first known cause of a majority of the outbreaks (@tbl-outbreaks-by-first-known-cause and @fig-outbreaks-by-first-known-cause).

Future research should be conducted to assess the state of disease outbreaks in long-term care homes in particular, both in Toronto and across Canada, in order to gain deeper insight into this critical issue. The prevalence of disease outbreaks in additional settings, such as schools and workplaces, should also be investigated to expand on this analysis.

@@
